### 2017-08-19 reports/thesis/book.tex
Saddlepoint in thesis.

%\frontmatter
\section*{Abstract}
X-ray CT can be used for defect detection and quality control of 3D printing. This report will outline statistical methods for comparing x-ray scans of a 3D printed sample with its computer model under the face of uncertainty. Shading correction methods and the mean and variance relationship were investigated. Lastly the compound Poisson was reviewed.

@@

%\mainmatter

@@

The p.d.f.~can be shown to be
\begin{equation}
p_X(x) = 
\delta(x) \euler^{-\lambda}
+
\sum_{y=1}^{\infty}\frac{\beta^{y\alpha}}{\Gamma(y\alpha)}x^{y\alpha-1}\euler^{-\beta x}\euler^{-\lambda}\frac{\lambda^y}{y!}
\quad\text{for }x\geqslant 0
\label{eq:compoundPoisson_pdf}
\end{equation}
and it is well known the infinite sum cannot be written in closed form. There are a number of approximations or computational methods to evaluate the p.d.f.~such as Fourier inverting the characteristic function \citep{dunn2008evaluation}, using the saddle point approximation \citep{daniels1954saddlepoint} or cleverly sum over certain terms in the infinite sum \citep{dunn2005series}. Of course direction simulation of a compound Poisson-Gamma random variable is incredibly easy if $U$ can be simulated. The p.d.f.~then can be estimated using these simulated samples.

@@

\chapter{Compound Poisson}
In this chapter, the compound Poisson-Gamma distribution is reviewed and studied. Firstly methods for evaluating the density of the compound Poisson-Gamma distribution are discussed. Approximations such as the Normal approximation or saddle point approximation \citep{daniels1954saddlepoint} can be used. These approximate densities can be written in closed form. Exact methods can be computationally challenging when dealing with the infinite sum but can be done when summing over important terms \citep{dunn2005series}. Next, a method for parameter estimation will be proposed here using the EM algorithm \citep{dempster1977maximum}. However in the end of the chapter, it will be shown that no unique maximum likelihood estimator exist when certain parameters of the compound Poisson-Gamma distribution gets too large.

@@

The saddle point approximation \citep{daniels1954saddlepoint} gives an approximate solution to inverting the Laplace transformation  of the m.g.f.~giving the p.d.f. However inverting the Fourier transformation of the characteristic function also gives the p.d.f.~using computational methods \citep{dunn2008evaluation}. The saddle point approximation will be studied here.

@@

Taking the Taylor's series of $M_U(\theta)$
\begin{equation}
M_U(\theta) = 1 + \sum_{r=1}^\infty\frac{\theta^r}{r!}M_U^{(r)}(0)
\quad\text{for }\lim_{r\rightarrow\infty}
\left\|
\frac{\theta}{r+1}\frac{M_U^{(r+1)}(0)}{M_U^{(r)}(0)}
\right\|<1
\end{equation}
and substituting it into $M_Z(\theta)$
\begin{equation*}
M_Z(\theta)=\exp\left[-\theta\frac{\sqrt{\lambda}\expectation[U]}{\sqrt{\expectation[U^2]}}\right]
\exp\left[
\lambda\sum_{r=1}^\infty\frac{\theta^r}{(\lambda\expectation[U^2])^{r/2}r!}M_U^{(r)}(0)
\right] \ .
\end{equation*}
Writing out in full the $r=1,2$ terms
\begin{multline*}
M_Z(\theta)=\exp\left[-\theta\frac{\sqrt{\lambda}\expectation[U]}{\sqrt{\expectation[U^2]}}\right]
\exp\left[
\theta\frac{\sqrt{\lambda}M_U'(0)}{\sqrt{\expectation[U^2]}}
+
\frac{\theta^2M_U''(0)}{2\expectation[U^2]}
\right.\\+\left.
\lambda\sum_{r=3}^\infty\frac{\theta^r}{(\lambda\expectation[U^2])^{r/2}r!}M_U^{(r)}(0)
\right] \ .
\end{multline*}
By using the fact that $\expectation[U^r]=M_U^{(r)}(0)$ for $r=0,1,2\dotdotdot$, a few terms cancel out to get
\begin{equation}
M_Z(\theta)=\exp\left[
\frac{\theta^2}{2}
+
\sum_{r=3}^\infty\frac{\theta^r \expectation[U^r]}{(\expectation[U^2])^{r/2}r!}\lambda^{1-r/2}
\right] \ .
\end{equation}

@@

>>>>

\frontmatter
\section*{Abstract}
X-ray CT can be used for defect detection and quality control of 3D printing. This report will outline statistical methods for comparing x-ray scans of a 3D printed sample with its computer model under the face of uncertainty. Shading correction methods and the mean and variance relationship were investigated. Lastly the compound Poisson was reviewed.

@@

\mainmatter

@@

The p.d.f.~can be shown to be
\begin{equation}
p_X(x) = 
\delta(x) \euler^{-\lambda}
+
\sum_{y=1}^{\infty}\frac{\beta^{y\alpha}}{\Gamma(y\alpha)}x^{y\alpha-1}\euler^{-\beta x}\euler^{-\lambda}\frac{\lambda^y}{y!}
\quad\text{for }x\geqslant 0
\label{eq:compoundPoisson_pdf}
\end{equation}
and it is well known the infinite sum cannot be written in closed form. There are a number of approximations or computational methods to evaluate the p.d.f.~such as Fourier inverting the characteristic function \citep{dunn2008evaluation}, using the saddlepoint approximation \citep{daniels1954saddlepoint} or cleverly sum over certain terms in the infinite sum \citep{dunn2005series}. Of course direction simulation of a compound Poisson-Gamma random variable is incredibly easy if $U$ can be simulated. The p.d.f.~then can be estimated using these simulated samples.

@@

\chapter{Compound Poisson}
In this chapter, the compound Poisson-Gamma distribution is reviewed and studied. Firstly methods for evaluating the density of the compound Poisson-Gamma distribution are discussed. Approximations such as the Normal approximation or saddlepoint approximation \citep{daniels1954saddlepoint} can be used. These approximate densities can be written in closed form. Exact methods can be computationally challenging when dealing with the infinite sum but can be done when summing over important terms \citep{dunn2005series}. Next, a method for parameter estimation will be proposed here using the EM algorithm \citep{dempster1977maximum}. However in the end of the chapter, it will be shown that no unique maximum likelihood estimator exist when certain parameters of the compound Poisson-Gamma distribution gets too large.

@@

The saddlepoint approximation \citep{daniels1954saddlepoint} gives an approximate solution to inverting the Laplace transformation  of the m.g.f.~giving the p.d.f. However inverting the Fourier transformation of the characteristic function also gives the p.d.f.~using computational methods \citep{dunn2008evaluation}. The saddlepoint approximation will be studied here.

@@

Taking the Maclaurin series of $M_U(\theta)$
\begin{equation}
M_U(\theta) = 1 + \sum_{r=1}^\infty\frac{\theta^r}{r!}M_U^{(r)}(0)
\quad\text{for }\lim_{r\rightarrow\infty}
\left\|
\frac{\theta}{r+1}\frac{M_U^{(r+1)}(0)}{M_U^{(r)}(0)}
\right\|<1
\end{equation}
and substituting it into $M_Z(\theta)$
\begin{equation*}
M_Z(\theta)=\exp\left[-\theta\frac{\sqrt{\lambda}\expectation[U]}{\sqrt{\expectation[U^2]}}\right]
\exp\left[
\lambda\sum_{r=1}^\infty\frac{\theta^r}{(\lambda\expectation[U^2])^{r/2}r!}M_U^{(r)}(0)
\right] \ .
\end{equation*}
Writing out in full the $r=1,2$ terms
\begin{multline*}
M_Z(\theta)=\exp\left[-\theta\frac{\sqrt{\lambda}\expectation[U]}{\sqrt{\expectation[U^2]}}\right]
\exp\left[
\theta\frac{\sqrt{\lambda}M_U'(0)}{\sqrt{\expectation[U^2]}}
+
\frac{\theta^2M_U''(0)}{2\expectation[U^2]}
\right.\\+\left.
\lambda\sum_{r=3}^\infty\frac{\theta^r}{(\lambda\expectation[U^2])^{r/2}r!}M_U^{(r)}(0)
\right] \ .
\end{multline*}
By using the fact that $\expectation[U^r]=M_U^{(r)}(0)$ for $r=0,1,2\dotdotdot$, a few terms cancel out to get
\begin{equation}
M_Z(\theta)=\exp\left[
\frac{\theta^2}{2}
+
\sum_{r=3}^\infty\frac{\theta^r \expectation[U^r]}{(\expectation[U^2])^{r/2}r!}\lambda^{1-r/2}
\right] \ .
\end{equation}

@@

\subsection{Saddlepoint approximation}
The m.g.f.~of a random variable $X$ is defined to be
\begin{equation}
M_X(\theta)=\expectation[\euler^{\theta X}]
=
\int_{-\infty}^{\infty}\euler^{x\theta} p_X(x) \diff x \ .
\end{equation}
For a given m.g.f., the saddlepoint approximation \cite{daniels1954saddlepoint} \cite{butler2007saddlepoint} finds an approximate $p_X(x)$. The saddlepoint approximation is given as
\begin{equation}
p_X(x)\approx\left(2\pi K_X''(s)\right)^{-1/2}\exp\left[K_X(s)-sx\right]
\label{eq:saddlePoint:generalSaddlePoint}
\end{equation}
where $K_X(\theta) = \ln\left(M_X(\theta)\right)$, and $s=s(x)$ is the solution to the saddle point equation $K_X'(s)=x$.

For the compound Poisson-Gamma distribution, the saddle point approximation is given as 
\begin{multline}
p_X(x)\approx
\frac{\left(\lambda\alpha\beta^\alpha\right)^{\frac{1}{2(\alpha+1)}}\euler^{-\lambda}}{\sqrt{2\pi(\alpha+1)}}x^{-\frac{\alpha+2}{2(\alpha+1)}}
\euler^{-x\beta}
\exp\left[x^{\frac{\alpha}{\alpha+1}}
	\frac{(\lambda\beta^\alpha)^{\frac{1}{\alpha+1}}(\alpha+1)}{\alpha^{\frac{\alpha}{\alpha+1}}}
\right]
\\
\text{for }x>0 \ .
\label{eq:saddle_point_approx}
\end{multline}
The integral of the density approximation over the support may not equal to one, it can be numerically re-normalised if necessary. Thus it may be more sensible to write the approximation up to a constant
\begin{equation}
p_X(x)\propto x^{-\frac{\alpha+2}{2(\alpha+1)}}\euler^{-x\beta}\exp\left[x^{\frac{\alpha}{\alpha+1}}
	\frac{(\lambda\beta^\alpha)^{\frac{1}{\alpha+1}}(\alpha+1)}{\alpha^{\frac{\alpha}{\alpha+1}}}
\right] \ .
\end{equation}

For the remainder of this section, it will be shown how the saddlepoint approximation can be algebraically derived. First of all the cumulant generating function $K_X(\theta)$ can be obtained from the m.g.f.
\begin{equation}
K_X(\theta) = \lambda\left[
\left(\frac{\beta}{\beta-\theta}\right)^\alpha-1
\right] \ .
\end{equation}
Taking the derivative with respect to $\theta$
\begin{equation}
K_X'(\theta)=\frac{\lambda\alpha\beta^\alpha}{(\beta-\theta)^{\alpha+1}}
\end{equation}
and this is known as the saddlepoint equation. The quantity $s=s(x)$ is the solution to the equation $K_X'(s)=x$, that is
\begin{equation*}
\frac{\lambda\alpha\beta^\alpha}{(\beta-s)^{\alpha+1}} = x
\end{equation*}
with solution
\begin{equation}
s = \beta - \left(\frac{\lambda\alpha\beta^\alpha}{x}\right)^{\frac{1}{\alpha+1}} \ .
\end{equation}

The second derivative of the cumulant generating function is
\begin{equation}
K_X''(\theta)=\frac{\lambda\alpha(\alpha+1)\beta^\alpha}{(\beta-\theta)^{\alpha+2}} \ .
\end{equation}
Substituting this and $K_X(\theta)$ into Equation \eqref{eq:saddlePoint:generalSaddlePoint}
\begin{equation*}
p_X(x)\approx
\frac{1}{\sqrt{2\pi}}\left[\frac{(\beta-s)^{\alpha+2}}{\lambda\alpha(\alpha+1)\beta^\alpha}\right]^{1/2}
\exp\left[
	\lambda\left(\left(\frac{\beta}{\beta-s}\right)^\alpha-1\right)-sx
\right] \ .
\end{equation*}
Substituting in $s=s(x)$
\begin{multline*}
p_X(x)\approx
\frac{1}{\sqrt{2\pi}}
\left[
\frac{\left(\beta-\beta+\left(\frac{\lambda\alpha\beta^\alpha}{x}\right)^{\frac{1}{\alpha+1}}\right)^{\alpha+2}}{\lambda\alpha(\alpha+1)\beta^\alpha}
\right]^{1/2}
\\
\exp\left[
	\lambda\left(
		\left(
			\frac{\beta}{\beta-\beta+\left(\frac{\lambda\alpha\beta^\alpha}{x}\right)^{\frac{1}{\alpha+1}}}
		\right)^\alpha
		-1
	\right)
	-x\left(
		\beta-\left(\frac{\lambda\alpha\beta^\alpha}{x}\right)^{\frac{1}{\alpha+1}}
	\right)
\right]
\end{multline*}
simplifying further
\begin{multline*}
p_X(x)\approx
\frac{1}{\sqrt{2\pi}}
\left[
\frac{\left(\frac{\lambda\alpha\beta^\alpha}{x}\right)^{\frac{\alpha+2}{\alpha+1}}}{\lambda\alpha(\alpha+1)\beta^\alpha}
\right]^{1/2}
\\
\exp\left[
	\lambda\left(
		\beta^\alpha
		\left(\frac{x}{\lambda\alpha\beta^\alpha}\right)^{\frac{\alpha}{\alpha+1}}
		-1
	\right)
	-x\beta
	+(\lambda\alpha\beta^\alpha)^{\frac{1}{\alpha+1}}x^{1-\frac{1}{\alpha+1}}
\right]
\end{multline*}
\begin{multline*}
p_X(x)\approx
\frac{1}{\sqrt{2\pi(\alpha+1)}}x^{-\frac{\alpha+2}{2(\alpha+1)}}
\left(\lambda\alpha\beta^\alpha\right)^{\left(\frac{\alpha+2}{\alpha+1}-1\right)/2}
\\
\exp\left[
	\lambda\left(
		\beta^\alpha
		\left(\frac{x}{\lambda\alpha\beta^\alpha}\right)^{\frac{\alpha}{\alpha+1}}
		-1
	\right)
	-x\beta
	+(\lambda\alpha\beta^\alpha)^{\frac{1}{\alpha+1}}x^{\frac{\alpha}{\alpha+1}}
\right]
\end{multline*}
\begin{multline*}
p_X(x)\approx
\frac{\left(\lambda\alpha\beta^\alpha\right)^{\frac{1}{2(\alpha+1)}}}{\sqrt{2\pi(\alpha+1)}}x^{-\frac{\alpha+2}{2(\alpha+1)}}
\\
\exp\left[
	-\lambda
	-x\beta
	+x^{\frac{\alpha}{\alpha+1}}
	\left(
		\frac{\lambda\beta^\alpha}{\left(\lambda\alpha\beta^\alpha\right)^{\frac{\alpha}{\alpha+1}}}+\left(\lambda\alpha\beta^\alpha\right)^{\frac{1}{\alpha+1}}
	\right)
\right]
\end{multline*}
\begin{multline*}
p_X(x)\approx
\frac{\left(\lambda\alpha\beta^\alpha\right)^{\frac{1}{2(\alpha+1)}}\euler^{-\lambda}}{\sqrt{2\pi(\alpha+1)}}x^{-\frac{\alpha+2}{2(\alpha+1)}}
\euler^{-x\beta}
\\
\exp\left[x^{\frac{\alpha}{\alpha+1}}
	\left(
		(\lambda\beta^\alpha)^{1-\frac{\alpha}{\alpha+1}}\alpha^{-\frac{\alpha}{\alpha+1}}
		+\left(\lambda\alpha\beta^\alpha\right)^{\frac{1}{\alpha+1}}
	\right)
\right]
\end{multline*}
\begin{multline*}
p_X(x)\approx
\frac{\left(\lambda\alpha\beta^\alpha\right)^{\frac{1}{2(\alpha+1)}}\euler^{-\lambda}}{\sqrt{2\pi(\alpha+1)}}x^{-\frac{\alpha+2}{2(\alpha+1)}}\euler^{-x\beta}
\\
\exp\left[x^{\frac{\alpha}{\alpha+1}}
	\left(
		(\lambda\beta^\alpha)^{\frac{1}{\alpha+1}}\alpha^{-\frac{\alpha}{\alpha+1}}
		+\left(\lambda\alpha\beta^\alpha\right)^{\frac{1}{\alpha+1}}
	\right)
\right]
\end{multline*}
\begin{multline*}
p_X(x)\approx
\frac{\left(\lambda\alpha\beta^\alpha\right)^{\frac{1}{2(\alpha+1)}}\euler^{-\lambda}}{\sqrt{2\pi(\alpha+1)}}x^{-\frac{\alpha+2}{2(\alpha+1)}}\euler^{-x\beta}
\\
\exp\left[x^{\frac{\alpha}{\alpha+1}}(\lambda\beta^\alpha)^{\frac{1}{\alpha+1}}
	\left(
		\alpha^{-\frac{\alpha}{\alpha+1}}+\alpha^{\frac{1}{\alpha+1}}
	\right)
\right] \ .
\end{multline*}

The expression $\alpha^{-\frac{\alpha}{\alpha+1}}+\alpha^{\frac{1}{\alpha+1}}$ can be simplified by putting the two terms over a common denominator
\begin{align}
\alpha^{-\frac{\alpha}{\alpha+1}}+\alpha^{\frac{1}{\alpha+1}} & = \alpha^{\frac{1}{\alpha+1}}+\frac{1}{\alpha^{\frac{\alpha}{\alpha+1}}}
\nonumber\\
& = \frac{\alpha^{\frac{1}{\alpha+1}}\alpha^{\frac{\alpha}{\alpha+1}}+1}{\alpha^{\frac{\alpha}{\alpha+1}}}
\nonumber\\
& = \frac{\alpha+1}{\alpha^{\frac{\alpha}{\alpha+1}}}
\end{align}
so that
\begin{equation*}
p_X(x)\approx
\frac{\left(\lambda\alpha\beta^\alpha\right)^{\frac{1}{2(\alpha+1)}}\euler^{-\lambda}}{\sqrt{2\pi(\alpha+1)}}x^{-\frac{\alpha+2}{2(\alpha+1)}}
\euler^{-x\beta}
\exp\left[x^{\frac{\alpha}{\alpha+1}}
	\frac{(\lambda\beta^\alpha)^{\frac{1}{\alpha+1}}(\alpha+1)}{\alpha^{\frac{\alpha}{\alpha+1}}}
\right] \ .
\end{equation*}
